---
title: "Wholesale Cust. Segmentation Using Hierarchical and K-Means Clustering"
author: "AhmedLubis"
date: "2026-07-30"
output: github_document
---

## Preparation ##
```{r}
#import_data
customer_data <- read.csv("wholesale_customers_data.csv")

head(customer_data)

#data_structure
str(customer_data)
summary(customer_data)

#data_preparation
data_cluster <- customer_data[, c(
  "Fresh",
  "Milk",
  "Grocery",
  "Frozen",
  "Detergents_Paper",
  "Delicassen"
)]

head(data_cluster)
```

## Exploration ##
```{r}
#scatterplot_matrix
pairs(data_cluster)

#correlation
cor(data_cluster)

#histogram
hist(data_cluster$Fresh,
     main = "Distribusi Fresh",
     xlab = "Fresh")
hist(data_cluster$Milk,
     main = "Distribusi Milk",
     xlab = "Milk")
hist(data_cluster$Grocery,
     main = "Distribusi Grocery",
     xlab = "Grocery")
hist(data_cluster$Frozen,
     main = "Distribusi Frozen",
     xlab = "Frozen")
hist(data_cluster$Detergents_Paper,
     main = "Distribusi Detergents_Paper",
     xlab = "Detergents_Paper")
hist(data_cluster$Delicassen,
     main = "Distribusi Delicassen",
     xlab = "Delicassen")
```

It can be seen that every histogram is right-skewed, so i have to transform
into log-form.

```{r}
#all_histogram_right-skewed
#copy_og_data
data_log <- data_cluster
#log_transformation-(log_base_e): column 1-6
data_log[, 1:6] <- log(data_cluster[, 1:6] + 1)
```

Exploration data_log
```{r}
#scatterplot_matrix
pairs(data_log)

#correlation
cor(data_log)

#histogram_log_form
#fresh
hist(data_log$Fresh,
     main = "Distribusi Fresh",
     xlab = "Fresh")
#milk
hist(data_log$Milk,
     main = "Distribusi Milk",
     xlab = "Milk")
#grocery
hist(data_log$Grocery,
     main = "Distribusi Grocery",
     xlab = "Grocery")
#frozen
hist(data_log$Frozen,
     main = "Distribusi Frozen",
     xlab = "Frozen")
#detergents_paper
hist(data_log$Detergents_Paper,
     main = "Distribusi Detergents_Paper",
     xlab = "Detergents_Paper")
#delicassen
hist(data_log$Delicassen,
     main = "Distribusi Delicasse",
     xlab = "Delicassen")
```

After exploration data, i have to do standardization data.

```{r}
#standardization_data
data_scaled <- scale(data_log)

head(data_scaled)
```

Next, the step should be done by me is measuring of distance.

```{r}
#distance_measurement
#euclidean_distance
euclidean_distance <- dist(
  data_scaled,
  method = "euclidean"
)

as.matrix(euclidean_distance)[1:5,1:5]
#manhattan_distance
manhattan_distance <- dist(
  data_scaled,
  method = "manhattan"
)

as.matrix(manhattan_distance)[1:5,1:5]
#chebyshev_distance
chebyshev_distance <- dist(
  data_scaled,
  method = "maximum"
)

as.matrix(chebyshev_distance)[1:5,1:5]
```

```{r}
#based_on_max_spare
minkowski_distance <- dist(
  data_scaled,
  method = "minkowski",
  p = 3
)

as.matrix(minkowski_distance)[1:5,1:5]
```

After measure of distance, I will do cluster analysis.

```{r}
#hierarchial_clustering
#complete_linkage
hc_complete <- hclust(
  euclidean_distance,
  method = "complete"
)

plot(
  hc_complete,
  hang = -1,
  main = "Dendrogram Complete Linkage"
)
```

```{r}
#creating_cluster
cluster_hc <- cutree(
  hc_complete,
  k = 5
)

cluster_hc
```

```{r}
#adding_cluster_to_data
customer_data$cluster_hc <- cluster_hc

head(customer_data)
```

```{r}
#cluster_visualization
library(factoextra)

fviz_cluster(
  list(
    data = data_scaled,
    cluster = cluster_hc
  )
)
```

```{r}
library(dplyr)

customer_data%>%
  group_by(cluster_hc) %>%
  summarise(
    Mean_fresh = mean(Fresh),
    Mean_milk = mean(Milk),
    Mean_grocery = mean(Grocery),
    Mean_frozen = mean(Frozen),
    Mean_detergents_paper = mean(Detergents_Paper),
    Mean_delicassen = mean(Delicassen)
  )
```

```{r}
#determine_optimal_number_of_cluster
#elbow_method
fviz_nbclust(
  data_scaled,
  kmeans,
  method = "wss"
)
#silhoutte_method
fviz_nbclust(
  data_scaled,
  kmeans,
  method = "silhouette"
)
```

Next step is k-means clustering. The "k" in k-means represents the number of 
clusters you want to find. The "means" refers to the average position of all
the data points within a cluster (the cluster center, or centroid).

```{r}
#k-means_clsutering
set.seed(123)

km <- kmeans(
  data_scaled,
  centers = 4,
  nstart = 25
)

km
```

```{r}
fviz_cluster(
  km,
  data = data_scaled
)
```

```{r}
#adding_the_result_of_cluster
customer_data$Cluster_KMeans <- km$cluster

head(customer_data)
```

```{r}
#cluster_evaluation
#calinski-harabasz_index
library(clusterSim)

index.G1(
  data_scaled,
  km$cluster
)
#davies-bouldin_index
index.DB(
  data_scaled,
  km$cluster
)$DB
```

```{r}
#comparison_clustering_model
# Complete linkage
hc_complete <- hclust(
  euclidean_distance,
  method = "complete"
)

# Average linkage
hc_average <- hclust(
  euclidean_distance,
  method = "average"
)

# Single linkage
hc_single <- hclust(
  euclidean_distance,
  method = "single"
)

# Ward linkage
hc_ward <- hclust(
  euclidean_distance,
  method = "ward.D2"
)
```

```{r}
#comparison_4clusters
cluster_complete <- cutree(hc_complete, k = 4)

cluster_average <- cutree(hc_average, k = 4)

cluster_single <- cutree(hc_single, k = 4)

cluster_ward <- cutree(hc_ward, k = 4)
```

Final step, I have to do some evaluations.

```{r}
#evaluated_by_silhouette_index
library(cluster)

sil_complete <- silhouette(
  cluster_complete,
  euclidean_distance
)

sil_average <- silhouette(
  cluster_average,
  euclidean_distance
)

sil_single <- silhouette(
  cluster_single,
  euclidean_distance
)

sil_ward <- silhouette(
  cluster_ward,
  euclidean_distance
)

mean_complete <- mean(sil_complete[,3])

mean_average <- mean(sil_average[,3])

mean_single <- mean(sil_single[,3])

mean_ward <- mean(sil_ward[,3])

mean_complete

mean_average

mean_single

mean_ward
```

```{r}
#evaluated_by_davies-bouldin_index
library(clusterSim)

db_complete <- index.DB(
  data_scaled,
  cluster_complete
)$DB

db_average <- index.DB(
  data_scaled,
  cluster_average
)$DB

db_single <- index.DB(
  data_scaled,
  cluster_single
)$DB

db_ward <- index.DB(
  data_scaled,
  cluster_ward
)$DB

db_complete

db_average

db_single

db_ward
```

```{r}
#evaluated_by_calinski-harabasz_index
ch_complete <- index.G1(
  data_scaled,
  cluster_complete
)

ch_average <- index.G1(
  data_scaled,
  cluster_average
)

ch_single <- index.G1(
  data_scaled,
  cluster_single
)

ch_ward <- index.G1(
  data_scaled,
  cluster_ward
)

ch_complete

ch_average

ch_single

ch_ward
```

```{r}
#resume_comparison_model
compared_result <- data.frame(
  Metode = c(
    "Complete",
    "Average",
    "Single",
    "Ward"
  ),

  Silhouette = c(
    mean_complete,
    mean_average,
    mean_single,
    mean_ward
  ),

  Davies_Bouldin = c(
    db_complete,
    db_average,
    db_single,
    db_ward
  ),

  Calinski_Harabasz = c(
    ch_complete,
    ch_average,
    ch_single,
    ch_ward
  )
)

compared_result
```

### **Determining the Optimal Number of Clusters**

Based on the evaluation results, the optimal clustering model is **K-Means with $k = 4$**. 

#### **Key Supporting Factors**

1. **Elbow & Silhouette Methods ($k = 4$ vs $k = 5$)**
   * Visual evaluation using `fviz_nbclust` showed cluster overlap at $k = 5$. 
   * Setting **$k = 4$** eliminated overlapping, yielding better-separated groups.

2. **Superior Cluster Separation (Calinski-Harabasz Index)**
   * K-Means ($k = 4$) achieved a **Calinski-Harabasz score of 131.36**.
   * Outperformed all tested hierarchical methods:
     * **K-Means ($k=4$):** `131.36` *(Highest)*
     * **Ward's:** `108.60`
     * **Complete:** `38.90`
     * **Average:** `8.30`

3. **Davies-Bouldin Index & Model Stability**
   * K-Means ($k = 4$) achieved a **Davies-Bouldin Index of 1.69**.
   * Although Average Linkage scored lower (0.35), it suffered from severe cluster imbalance/chaining effect.

4. **Balanced Cluster Distribution**
   * K-Means ($k = 4$) produced well-balanced cluster sizes: **130, 61, 119, and 130**.
   * Avoided single-cluster dominance seen in Average/Single linkage, making the segments practical for business interpretation.

---

> **Conclusion:** **K-Means ($k = 4$)** is selected as the optimal model because it provides the highest internal validity score (Calinski-Harabasz) and the most stable, balanced cluster structure.